|

### 2020-10-23 paper/v1/broadband.speed.covid.Rmd
Intro drafted. Where says 'insert ref' I think I have some ideas, but if I've added a question mark, I'm not sure. Also, some bits copied across into data and results sections. Needs more work and discussion, but hopefully captures many of our ideas discussed yesterday. I may have more time before you do, in which case I will commit further changes.

[our Google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9)

**PARA1**: working from home is a challenge and maybe an opportunity [use arguments from @BUDNITZ2020102713]
It has been altered dramatically because of covid

**PARA2**: covind and working from home, cities, urban structure
Potentially useful readings for covid and cities:

@@

It is important to become aware of working from home patterns because of:
- transport planning reasons (see para 1)
- WFM might increase coparing to pre-covid level

**PARA3**: How can we observe observe WFM? 
Directly through surveys (@felstead2020homeworking).
Add survey limitations (cost, reprentativeness, time lag between survey and data availability)
Data gap.
Passive data collection through internet speeds
Another option could be mobility data.
How is internet speed data a better source than mobility?

**PARA4**: Contention

**PARA5**: Data and methods

**PARA6**: Contribution:

## broadband studies, divides, broadband tech stuff

## from telecommuting to #WFH

@@

Data details and some figures, descriptive stats

@@

Clusters, cluster description and aux regressions

>>>>

**PARA1**During the pandemic, working from home using Information and Communication Technologies (ICT), whether partially or exclusively, was transformed from a niche means of accessing work, albeit one that had been on a slow, upward trend, to a widespread way of life in many countries (Felstead and Reuschke, 2020; **insert further refs from elsewhere**). The ability to work from home or telecommute has enabled millions to retain their jobs and, to a varying extent, maintain productivity during periods of strict lockdown around the world. However, this ability has not been evenly distributed socially or spatially, creating a new type of digital divide. On one side are those who can work from home and have been able to enjoy both economic resilience and greater personal safety. On the other side, previously employed individuals have been forced to accept furlough or redundancy packages unless they are part of the cadre of essential workers, who are potentially at high risk of infection. Using the UK as a case study, this paper aims to improve our understanding of the spatial and social dimensions of this new digital divide by assessing the resilience of broadband speeds in terms of both quality and reliability of service, and whether this reinforces or redresses prior digital divisions.

**PARA2**The capability to work from home has previously been studied from the perspective of whether work tasks in a given occupation both can be and are allowed to be performed independently of location or co-location with colleagues, including supervisors (**insert refs**). However, successful telecommuting also requires that the quality and reliability of ICT services, particularly home internet connection speeds, enable the completion of work tasks with a minimum of delay or interruption. In reality, the performance of broadband speeds with respect to telecommuters has never been tested before at scale, as working from home and connecting to colleagues and workplace resources via broadband has previously been the purview of a small minority of workers. In the UK, the most recent statistics indicate that **x%** of working age adults identified their home as their usual workplace, with an additional **x%** who said they work from home at least once a week (**insert stats from 2018 or 2019 NTS - also x% of these use ICT to complete their work at home**). Their demands on internet services have been minor compared to the demands of leisure users, such that the broadband performance offered by different Internet Service Providers (ISPs) has been benchmarked according to download speeds during the evening 'primetime', when video streaming services are at their peak, rather than during the working  day (**OfCom, 2017**). Yet the pandemic has fundamentally changed not only how many people work from home - 47% of working age adults at the peak in April, 2020 (**ref**) - but also their technical requirements when doing so. Whilst it may have been common in the past for telecommuters to complete solitary work tasks at home, but still visit their workplace for meetings (**ref?**), during the pandemic the replacement of face-to-face contact with bandwidth-intensive video conference calling was seen as essential to suppressing the spread of infection. This type of broadband use can particularly affect upload speeds **ref?**, so we seek to develop a new benchmark (**is it a benchmark?**)to assess the quality and reliability of upload speeds by time of day and day of the week across local authorities in the UK during the time when the population were told to work from home if at all possible.

**PARA3**The experience of upload speeds is not the same as the maximum speeds offered by an ISP or possible speeds over a particular type and length of connection. Therefore, we use volunteered geographic information in the form of speed-checks run by users to test their experienced broadband speeds, upload and download. Meanwhile, the quality and reliability of upload speeds, like ICT more generally, vary in time and space due to both supply and demand-side influences, and can be measured in a number of ways. These include: a) mean, experienced connection speed, b) standard deviation or the amount of fluctuation from the mean, and c) the variation in speeds at particular times of day when working from home is more likely to take place. We take account of all three measurements in order to describe upload speeds as fully as possible. We start by calculating the mean upload speed for every local government district in the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank holidays. We then cluster the local authorities by these temporal profiles, allowing us to identify patterns and describe the overall means, standard deviations, and other relevant statistics for each spatial cluster. Thus we aim to answer the first part of our research question: How resilient are broadband speeds as experienced in different parts of the UK during a time of extreme demand?

**PARA4**The cause of these different experiences of broadband resilience may be different in different areas, as they may reflect similarities in patterns of demand or similar quality of infrastructure. Our approach is also limited by potential endogeneity, as for example, better quality connections with high mean speeds may enable more working from home, but greater demand may cause slower speeds, less reliability and greater variability of speed at different times of day or week. Therefore, we avoid attributing any cause to our analysis of the experienced level of quality and reliability of upload speeds. Instead, we run auxilliary regressions in order to understand how the spatial and temporal patterns of internet service relate to the economic geography of the UK. We discuss how the different patterns might support or undermine efforts to work from home and maintain safe productivity and whether they reinforce existing spatial and social inequalities. From this analysis, we hope to provide a greater understanding of how telecommuting and technology intersect at a time of extreme demand, and what lessons this time has for a future where telecommuting is likely to remain a common means of accessing work and broadband services, as well as infrastructure, must be fit for purpose.

**PARA5*The structure of this paper is as follows.

## broadband studies, digital divides, broadband tech stuff - benchmarking?

## from telecommuting to #WFH - Introduce space-time geography - circumstances where work can be carried out more flexibly in space and time- how relate to digital divides - capability theory of mobility? suppressed demand...

##covid and working from home, cities, urban structure - economic geography of UK
Potentially useful readings for covid and cities:

@@

@@

The primary data analysed in this paper was provided by Speedchecker Ltd, a private company that allows internet users to check their own broadband upload / download speeds, and stores every speed-check with a timestamp and geolocation.

Our approach involves aggregating all the speed-checks during the 13 weeks of March to May inclusive for weekdays in 2020 by each hour of the day and day of the week. As our research aims to identify the geography of internet service resilience for work purposes, bank holidays and the hours between midnight and 6am were also excluded. The composite week time series thus comprise 18 hours multiplied by 5 weekdays or 90 time points per series. These time series were calculated for each of the 382 Local Authority Districts (LADs) in the UK, standardised, and then a k-means clustering algorithm was applied, including dynamic time warping. The LADs were assigned to 10 clusters for upload speeds. The cluster id for each LAD is then reattached to the speed-check dataframe to identify the characteristics of each cluster, including number of LADs, and the descriptives statistics of upload speeds in that cluster, and the temporal profile by hour of the day and day of the week.

**Data details and some figures, descriptive stats - include whole sample time profile for 2019 and 2020 frequency of tests run as part of why we chose to create the time profiles by hour of the day and day of the week rather than daily over the whole period.**

@@

##Upload Clusters / cluster description

The temporal profiles used to cluster the local authorities have been summarised in graph [.], which shows a composite profile of mean upload speeds per hour per day for each cluster. For upload speeds, 345 of 382 local authorities, or over 62 million people, fall into cluster 6 or cluster 9. Graph [.] shows that both of these clusters have relatively similar temporal profiles, which are flatter than the other, smaller clusters, suggesting better reliability of service. However, the upload speeds at all times for cluster 9 are substantially higher than for cluster 6, which is an indication of better quality of service. This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59, as well as the evening peak period from 19:00-20:59. In comparison, the time profile in graph [.] shows upload speeds in cluster 1 are on average lower at certain times of day during the study period than any other cluster, whilst the profile for cluster 3 appears to show speeds fluctuating as much as cluster 1, but at levels usually higher even than cluster 9.

The variability of the smaller clusters may be related to the fewer speed tests from fewer local authorities that have been averaged, whilst averaging greater numbers of speed tests could artificially flatten the profile. This appears not to be the case for cluster 9, as there is negligible difference between morning and evening upload speeds, at 1% slower in the morning, confirming a high level of reliability of service. In comparison, upload speeds in cluster 6 are 4% slower in the morning than in the evening and experienced the joint highest ratio of standard deviation to mean across the time period under assessment. This suggests that although the time profile is relatively flat in graph [.], the experience is one of speeds that fluctuate from a lower mean, and therefore might more often impact on online activities. Still, in most of the smaller clusters the reliability of service is worse than in cluster 9 in terms of the ratio of standard deviation to mean and worse than cluster 6 as well in terms of the ratio of upload speed in the morning peak compared to the evening peak. In five of the smaller clusters, which are home to almost 3 million people, speeds are 12-19% slower between 9:00-10:59 compared to 19:00-20:59.

The quality and reliability of broadband service is thus much better in the 115 Local Authority areas in cluster 9, which are mostly in urban or suburban areas, compared to cluster 6. These include 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation. There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, and many suburban districts from the South East of England to South Tyneside. Meanwhile, the 230 local authorities in cluster 6, which have lower speeds on average and more variation in service still include major urban areas, such as Bristol, Liverpool and Leeds, and many suburban areas, but also include some of the most rural areas in the country. Meanwhile, Cluster 1, with 10 local authorities that are home to over 1 million people has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean. This cluster's most populous area is Westminster in central London.

**further description of temporal profiles of other clusters - include all in graph?** Indeed, the only cluster where upload speeds were slower in the evening than in the morning was cluster 5, made up of 2 local authorities with less than 200,000 people: Three Rivers, a suburban district north of London, and Fylde, a seaside suburb of Blackpool. However, these are likely to be outliers and may not have many tests from which the clusters are calcualted. **Take out smaller clusters?** The exceptions can be found in seven of the eight much smaller clusters including 35 local authority districts, where AM peak upload speeds are between 6% and 18% slower than PM peak upload speeds, although the mean speeds for each cluster are higher than cluster 6. Indeed, in 25 local authorities with a combined population of almost 3 million, speeds are 13% or more slower in the morning than in the evening. Included in this latter group are central London borough of Westminster and the London Borough of Newham, rural authorities like Eden and West Devon, and small cities like Dundee and Carlisle.

##aux regressions

Auxiliary regressions indicate that the speed tests in cluster 9 authorities are also more likely to have been run on services provided by Virgin Media, suggesting they are in the half of the country with the most lucrative ICT market, which originally attracted the cable TV provider (OfCom...). For example, although auxiliary regressions show that Cluster 6 local authorities are more likely to be in the South of the UK than Cluster 9, the cluster notably includes Southern rural districts from Cornwall to North Norfolk. Slower speeds could reflect the lower quality of service in rural areas compared to urban and suburban areas. There is frustration, however, as auxiliary regressions show that those living in cluster 6 had the highest probability of testing their broadband between 9:00 and 11:00 of any of the 10 upload speed clusters.

the finding in the auxiliary regressions that those in cluster 9 ran the fewest speed tests per person during the morning period of any cluster. Now this may be an indication of fewer people working from home, less contention, and less resultant frustration. Cluster 9 comprises many central urban areas and has the lowest number of established businesses per inhabitant, which could be interpreted as a dominance of large employers. However, the job density is lower than cluster 6, meaning there are not as many jobs per resident in these areas. As the cluster includes many suburban areas too, which may be largely residential, could the quality and reliability of internet service be reinforcing patterns of telecommuting by those in wealthier suburbs who can work from home? Earnings in cluster 9 are second highest of all the clusters, with only Cluster 2 (comprising just North East Lincolnshire and East Lothian, population 265k) earning more per person.

The auxiliary regressions show that compared to the two authorities in cluster 5, all the other clusters had a lower percentage of working people in managerial, professional and administrative jobs.??? Yet the auxiliary regression suggests that there are not many tests being run during the am peak in cluster 1. This may be because there are fewer people working at home checking their broadband than in most other clusters. Indeed, the auxiliary regressions indicate that cluster 1 has the highest job density or proportion of jobs to working-age population, which is likely to due to the presence of Westminster, central London, cluster 1's most populous local authority. Westminster not only has more workplaces than residents, but it is reasonable to presume that many who would normally work in Westminster, but be able to work from home during lockdown are likely to live outside central London and not be subject to the fluctuating speeds there. Workplaces, meanwhile, some of which would still have been open, could be running programmes that cause the slowdown and variation, but would be more likely to have their own in-house diagnostics, rather than using a service like Speedchecker Ltd.
